---
title: "lab_02"
author: "derek willis"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
#install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: <https://worldpopulationreview.com/us-counties/maryland>.

**A1**
There appear to be 17,083 new registrants since July 21. Montgomery County, Baltimore City, and Baltimore County have seen the most at 4848, 2801, and 1683 new registrants respectively. Montgomery County is the most populous county in Maryland. Baltimore City and Baltimore County are both in the top 5 most populous counties in Maryland.

```{r}
new_voters<-md_voters_2024 |>
filter(STATE_REGISTRATION_DATE>"2024-07-21") 

```


```{r}
new_voters |>
group_by(COUNTY) |>
summarise(
   county_amount = n()
   ) |>
  arrange(desc(county_amount))
```


**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**A2**
The number of new registrations generally increases over time with some variation.In the range between July 21 and August 7 (weeks 29-32) there is a significant jump, from 3906 registrations per week in week 29 to 7164 per week in week 30. Those numbers are higher than the average number of registrations per week, 3451.125, but the highest number of registrations was actually in week 24. 

```{r}
md_voters_2024 <-md_voters_2024|>
  mutate(
  registration_week = week(COUNTY_REGISTRATION_DATE)) 
```

```{r}
md_voters_2024|>
group_by(registration_week) |>
  summarise(
   number_of_registrations = n()) |>

arrange(registration_week)
  
  
```

```{r}
md_voters_2024|>
  group_by(registration_week) |>
  summarise(
    number = n())|>
  summarise(
  average_number_of_registrations = mean(number), 
  min_amount = min(number),
  max_amount = max(number))
  
```


**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2? 

**A3** The largest number of registrations were in July, and the next largest number was in June. This might suggest an increase in registrations during the ramp up to the election that was further boosted when Kamala Harris became the democratic nominee. 

```{r}
md_voters_2024|>
  mutate(
  registration_month = month(COUNTY_REGISTRATION_DATE))|>
  group_by(registration_month) |>
  summarise(
    number = n())
  

```

**Q4** Maryland has, officially, five regions: <https://www.visitmaryland.org/article/maryland-regions>. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**A4** The central and capital regions account for the most new voters at 7587 and 7202	respectively. The democratic party accounts for the most new voters at 7644. The second largest group of new voters (6052) are registered as "unaffiliated." 

```{r}
md_voters_2024 <-md_voters_2024|>
  mutate(
   region = case_when(
  COUNTY == "Garrett" ~ "Western", 
  COUNTY == "Washington" ~ "Western",
  COUNTY == "Allegany" ~ "Western", 
  COUNTY == "Frederick" ~ "Capital",
  COUNTY == "Montgomery" ~ "Capital",
  COUNTY == "Prince George's" ~ "Capital",
  COUNTY == "Calvert" ~ "Southern", 
  COUNTY == "Charles" ~ "Southern", 
  COUNTY == "Saint Mary's" ~ "Southern", 
  COUNTY == "Kent" ~ "Eastern Shore", 
  COUNTY == "Queen Anne's" ~ "Eastern Shore", 
  COUNTY == "Talbot" ~ "Eastern Shore", 
  COUNTY == "Caroline" ~ "Eastern Shore", 
  COUNTY == "Dorchester" ~ "Eastern Shore", 
  COUNTY == "Wicomico" ~ "Eastern Shore", 
  COUNTY == "Somerset" ~ "Eastern Shore", 
  COUNTY == "Worcester" ~ "Eastern Shore", 
  COUNTY == "Cecil" ~ "Eastern Shore", 
  COUNTY == "Anne Arundel" ~ "Central", 
  COUNTY == "Baltimore City" ~ "Central", 
  COUNTY == "Baltimore County" ~ "Central", 
  COUNTY == "Carroll" ~ "Central", 
  COUNTY == "Harford" ~ "Central", 
  COUNTY == "Howard" ~ "Central"
))
```

```{r}
new_voters <-new_voters |>
  mutate(
   region = case_when(
  COUNTY == "Garrett" ~ "Western", 
  COUNTY == "Washington" ~ "Western",
  COUNTY == "Allegany" ~ "Western", 
  COUNTY == "Frederick" ~ "Capital",
  COUNTY == "Montgomery" ~ "Capital",
  COUNTY == "Prince George's" ~ "Capital",
  COUNTY == "Calvert" ~ "Southern", 
  COUNTY == "Charles" ~ "Southern", 
  COUNTY == "Saint Mary's" ~ "Southern", 
  COUNTY == "Kent" ~ "Eastern Shore", 
  COUNTY == "Queen Anne's" ~ "Eastern Shore", 
  COUNTY == "Talbot" ~ "Eastern Shore", 
  COUNTY == "Caroline" ~ "Eastern Shore", 
  COUNTY == "Dorchester" ~ "Eastern Shore", 
  COUNTY == "Wicomico" ~ "Eastern Shore", 
  COUNTY == "Somerset" ~ "Eastern Shore", 
  COUNTY == "Worcester" ~ "Eastern Shore", 
  COUNTY == "Cecil" ~ "Eastern Shore", 
  COUNTY == "Anne Arundel" ~ "Central", 
  COUNTY == "Baltimore City" ~ "Central", 
  COUNTY == "Baltimore County" ~ "Central", 
  COUNTY == "Carroll" ~ "Central", 
  COUNTY == "Harford" ~ "Central", 
  COUNTY == "Howard" ~ "Central"
))
```


```{r}
new_voters |>
  group_by(region) |>
  summarise(
    number = n()) |>
  arrange(desc(number))
  
```
```{r}
new_voters |> 
   group_by(PARTY) |>
  summarise(
    number = n()) |>
  arrange(desc(number))
```


**Q5** What's an additional question you could ask of this data?


**A5** What residential city accounts for the most new voters?
In each region of Maryland, what's the breakdown of new voters by political party? 
