---
title: "Class Reference"
author: "Haley Parsley"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---
08/30/2024: Some Notes on Basic R Commands and Functions

Maybe the most basic command...
```{r}
2+2
```

```
Use the left facing arrow to assign things to variables. And use * for multiplication.

```
```{r}
first <-1
second <-2
second * first 
```

```
You can have as many variables as you want. And you can name them whatever you want. 

```
```{r}

butterfly <-3
spider <- 6
ant <- 10
butterfly * spider * ant
```

```
About Functions


HINT: "A function is a little bit of computer code that takes in information and follows a series of pre-determined steps and spits it back out. A recipe to make pizza is a kind of function. We might call it make_pizza()." In this example, make_pizza take the ingredients of a pizza, passes them through preset outlined steps (kneading, baking, etc) to make the output of make pizza(), a finished pie. Also -- a good argument for this function might be make pizza (pepperoni, cheese, flour, water, oil, salt, etc) for output: pepperoni pizza.

'''

Eventually, we will use various libraries to help us analyze data.

Here are some functions we will use when we analyze campaign contribution data (in a CSV file format) using tidyverse libraries:

glimpse (variable name) gives a list of the columns, data type of each column, and first few values of each column. 

head(variable name) prints out the column and first six rows of data.

Some other functions...
in the dplyr library, the summarize (variable name) function counts things and creates a new data frame (I'm not sure what this means)

arrange (variable name) arranges data in ascending order. If written like this, it will sort in descending order: arrange(desc(count_contribs))

```{r}

```

```


 group_by(date)
  summarise()
  arrange(desc(count_contribs))

lubridate() corrects the month/day/year order of dates in CSV
head() shows top six lines
colnames() shows the row and column names 
glimpse() shows rows and columns
 mutate(
 percent_election_day = (election_day/votes)*100
)            use *100 to make the result a percentage

mutate makes a new column.

  arrange(desc(percent_election_day)). arrange desc. orders from largest to smallest
  
  
  > week('2024-07-25')
[1] 30

guess_max=10 guess the data type of each column using first 10 rows
WHEN USING GUESS/problem, if it finds one even character that's not a number, it will automatically recognize that column's values as characters 

.default = (nameofcolumn) at the end of function series ...MEANS don't change anything but what the function said. See example ->
texas_precinct_20 <- texas_precinct_20 |>
  mutate(election_day = case_when(
    election_day == '5+' ~ '49',
    .default = election_day
  ))
  
  ----
 use as.numeric to say that everything in a column is a number (make sure that everything can be run as a number before running this) example ->
 texas_precinct_20 <- texas_precinct_20 |> mutate(election_day = as.numeric(election_day))


Janitor makes header rows consistent in terms of case, where _ (gets rid of spaces) are inserted, etc. --> it's the clean_names(function) and it's just for names of columns.  

NO COLUMN NAMES THAT BEGIN WITH NUMBERS!!!!! Janitor will put an x in front to fix this.






































```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2024_fall")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```
