---
title: "lab_04"
author: "Derek Willis"
date: "2024-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(janitor)
library(lubridate)

```

For this lab, we want to investigate spending by the two leading Senate candidates in Maryland during this election, Angela Alsobrooks and Larry Hogan. Our goal is to try and standardize the purpose of the expenditures so that we can compare the two campaigns. We have the data, but it's messy - purpose descriptions are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of campaign expenditures located in the data folder called "md_senate_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
campaign_expenses <- read_csv("data/md_senate_expenses.csv")

```

## Answer questions

**Q1.** You've been assigned to compare the spending priorities of the two campaigns, but they describe their spending differently using the `purpose` column. Using OpenRefine, create a project using the original CSV file, make a copy of the `purpose` column called `purpose_clean` and then standardize it, focusing on the purposes that appear most often or have the largest total amounts. You can transform the data as you want - making all records the same case, for example, to help. The goal is to be able to group together as many similar purpose values, so you can choose to call all transportation-related expenses the same thing.

How do you know when you're done? It's a judgment call, but you should be close to 100 unique values for `purpose_clean`, and maybe even below that number.

Then, when you are done standardizing `purpose_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.

Then, using that `purpose_clean` column, write code to find the largest amounts of spending on categories by both campaigns.

**A1.** Of all its spending categories, Hogan's campaign spent the most money on Media and Ads ($2,357,161.67), Payroll ($502,604.49), Fundraising ($315,492.06), Printing ($199014.36) and Strategy Consulting ($124,706.00).

Of all its spending categories, Alsobrook's campaign spent the most money on Media and Ads ($4,553,131.09), Payroll ($825,083.08), List Acquisition	($529,936.97), and Payroll Taxes ($436,720.02).	

```{r}
clean_campaign_expenses <- read_csv("data/clean_md_senate_expenses.csv")
```


```{r}
clean_campaign_expenses |>
  filter(candidate == "Hogan") |>
  group_by(candidate, Purpose_clean_grouped) |> 
  summarize(total_amount = sum(amount)) |> 
  arrange(desc(total_amount)) 
```
```{r}
clean_campaign_expenses |>
  filter(candidate == "Alsobrooks") |>
  group_by(candidate, Purpose_clean_grouped) |> 
  summarize(total_amount = sum(amount)) |> 
  arrange(desc(total_amount)) 
```


**Q2.** Let's do some research on the top purposes from A1. Feel free to search the Web to understand those terms better if need be, and then write a description comparing and contrasting the two campaigns' largest categories of spending.


**A2.**
Alsobrooks paid 529,936.97 for list acquisition, meaning the purchase of donor names and contact info from, in this case, Break Something, Inc., a campaign consulting firm. Hogan hasn't listed any list acquisition expenses. This could be because he has these lists from prior campaigns. Or, he could have grouped this expenditure category into another when dislcosing it-- maybe his "strategy consulting," for example (this could probably be true of many expenditures by both campaigns -- it's hard to know if we're really making a 1:1 comparison, especially in instances where the data bas been further grouped using OpenRefine). As another example, Alsobrooks has a separate expenditure listing for "payroll taxes," and it's quite high. I would expect that Hogan similarly has to may these taxes because he has payroll expenses, but he hasn't disclosed any payroll tax expenditures-- so did Hogan group these taxes into another category?

Both candidates spend a lot of money on credit card and bank fees (processing donations), as well as on payroll and fundraising costs.

**Q3.** Pick one of the top categories for each campaign and write code that shows how that spending has changed over time, if at all. You'll need to use lubridate for this. Write a couple of sentences describing that pattern.

**A3.** Overall, Alsobrooks has spent much more money than Hogan on media and ads. She started spending money on ads in mid-2023, while Hogan didn't begin spending on ads until March. Hogan spent the most money on media and ads ($805,732.85) during week 21 of 2024 (late May). Alsobrooks spent the most on media and ads ($605,060.00) during week 18 of 2024 (late April and early May) --right before the 2024 primary election on May 14. In the weeks leading up to the primary, Alsobrook's ads and media expenditures steadily increased. Hogan's spending has been high in the periods right before and right after the election, but his spending follows less of a clear pattern than Alsobrooks. It also appears that Hogan made his last ad and media expenditure in week 26 of 2024 (late June).

```{r}
weekly_expenses <- clean_campaign_expenses |>
  mutate(
  expenditure_week = week(date))|>
  mutate(
  expenditure_year = year(date))

```

```{r}
weekly_expenses |>
  filter(candidate == "Hogan")|>
  filter(Purpose_clean_grouped == "Media and Ads") |>
  group_by(expenditure_week, expenditure_year) |>
  summarise(total_amount_spent = sum(amount)) |>
  arrange(desc(total_amount_spent))


```


```{r}
weekly_expenses |>
  filter(candidate == "Alsobrooks")|>
  filter(Purpose_clean_grouped == "Media and Ads") |>
  group_by(expenditure_week, expenditure_year) |>
  summarise(total_amount_spent = sum(amount)) |>
  arrange(desc(total_amount_spent))


```





**Q4.** Which committee has the larger payroll or salary spending so far? How many unique recipients of payroll/staff spending are there for each candidate?

**A4.** Alsobrooks has spent the most on payroll ($825,083.10). Hogan has spent $502604.50. Alsobrooks has 24 unique recipients for payroll expenditures, while Hogan has three.

```{r}
clean_campaign_expenses |>
  filter(Purpose_clean_grouped == "Payroll") |>
  group_by(candidate) |> 
  summarize(total_amount_spent=sum(amount)) 
```

```{r}
clean_campaign_expenses |>
  filter(Purpose_clean_grouped == "Payroll") |>
  group_by(candidate) |> 
  summarize(total_payments = n()) 
```

```{r}
clean_campaign_expenses |>
  filter(Purpose_clean_grouped == "Payroll") |>
  group_by(candidate) |>
  summarize(
  payroll_recipients=n_distinct(last_name)) 

```

**Q5** What's the most interesting story idea you've found in this data?

**A5**
Alsobrook's ad spending peaked in the weeks leading up to the primary election, while Hogan's spending has been highest in the weeks following the primaries. What does this say about each candidate's campaign strategy? 

Also, it appears that Hogan does not list the specific members of his payroll, but instead lists the recipients of most payroll expenses as "STAFFING - SEE MEMO." Where is this memo?
